### 2014-01-28 inst/doc/pubs/one-simple-way.Rmd
writing on one simple way

@@

Use `EML`.

@@

White et al. outline "Nine simple ways to make it easier to (re)use your data".  Here we present one simple way to meet these criteria.  Before we worry about what the criteria are, let's just jump in and make our archival data shared:

@@

```{r, results="asis"}
dat = data.frame(river=c("SAC", "SAC", "AM"),
                 spp = c("king", "king", "ccho"),
                 stg = c("smolt", "parr", "smolt"),
                 ct =  c(293L, 410L, 210L))

@@

Define the units used in each column, or the abbreviations we use in factors.

@@

Specify the basic file metadata of author/contact, title, and a description, and choose where to publish:

```{r}
eml_write(dat, col_metadata, unit_metadata, 
          .contact="Carl Boettiger <cboettig@gmail.com>",  
          .title="A test EML file", 
          file_description="Test data, only intended for testing",
          destination="figshare")
```

@@

Sharing is as easy as `publish_EML("my_eml_data.xml")`

@@

`EML` can automatically push our data to a variety of established repositories, as illustrated in this example.

@@

We see by default `EML` has provided us with the liberal CC0 license for the data, consistent with major repositories such as Dryad.  Other options can be specified if needed.

Ethan White et al have recently published an important paper on "Nine simple ways to share your data", which has garnered considerable attention even as a preprint [^1]

[^1]: Disclaimer: I reviewed this paper, Karthik Ram served as handling editor.

>>>>

_Ideas in Ecology and Evolution_ recently published a paper by White _et
al._ that should be considered required reading in any graduate training
program Ecology or Evolution training program. Entitled _"Nine simple ways
to make it easier to (re)use your data"_, the authors present a thorough
and accessible overview of practices essential for any researcher but
getting increasing intention in light of spreading requirements from
primary publishers and funders of ecological and evolutionary research
for the formal deposition and archiving of data upon publication. The
spread of such requirements brings new urgency to the elements described
by White and colleagues, which echo sentiments and recommendations that
have appeared regularly in this literature over the previous years and
decades.  The development of standards such as the Ecological Metadata
Language, the emergence of suitable online data repositories and the
widespread adoption of scripting languages such as R among students of
Ecology and Evolution have all reduced barriers to sharing and reusing
data efficiently and effectively.

Here we introduce the `EML` package for the R statistical environment, 
as one simple way to accomplish these nine elements and much more besides.

<!--
These developments, together with new semantic technologies currently on
the horizon, and the billons of dollars invested in enormous monitoring
and data collection efforts such as the National Ecological Observatory
Network, the Ocean Observatory Intiative, and similar efforts offer
unpresidented opportunities to tackle the rich, integrative and data-intensive 
questions that challenge our globe today.  
-->

@@

@@

Before we worry about what the criteria are, let's just jump in and make our archival data shared:

@@

```{r, results="asis"}
dat = data.frame(river=c("SAC", "SAC", "AM"),
                 spp = c("king", "king", "ccho"),
                 stg = c("smolt", "parr", "smolt"),
                 ct =  c(293, 410, 210L))

@@

Define the units used in each column.  In the case of factors, we define
the abbreviations in a named string.

@@

```{r}
dat_with_metadata <- data.set(dat, col.defs = col_metadata, unit.defs = unit_metadata)
```

Specify the basic file metadata of creator and contact person, give it
a title, and even choose where to publish:

```{r}
eml_write(dat_with_metadata,
          contact = "Carl Boettiger <cboettig@ropensci.org>",  
          title = "A test EML file", 
          destination = "figshare")
```

Just like that, we have accomplished all nine steps, and more besides.
In the sections below, we'll unpack what this function has done, and how
to take advantage of this metadata when (re)using this data in the future.

<!--

If we had used just the original `data.frame` as the first argument, the
function would prompt us for the appropriate unit and column metadata. If
no destination is given, a filename can be provided instead.

@@

Sharing is as easy as `publish_eml("my_eml_data.xml")`

@@

The `EML` package can automatically push our data to a variety of
established scientific repositories, including some of those discussed by
White _et al._.  In the introductory example we submitted the data file to
figshare, an internationally recognized interdisciplinary data repository
that is integrated with PLOS and several Nature Publishing Group journals.
This command uploaded both the raw data in the recommended plain-text
CSV format and the XML file containing the EML metadata description to
the same figshare fileset object.  Like most repositories, figshare
requires basic metadata of it's own to enable it users to search and
discover content.  The `EML` package has automatically extracted the
dataset creators and title, as well as any description or keywords already
available in the metadata and copied them over to appropriate fields
in the figshare metadata. In return, figshare has supplied us with a
permanent identifier (DOI) that can be used to persistently locate and
cite this data, even if figshare itself goes out of business.

Other repositories may be even better suited for publishing of
EML-annotated data. The Knowledge Network for Biocomplexity (KNB) is
particularly appropriate for EML annotated data, as a member of the
DataONE network (see @Reichman2011) and because the repository already
"speaks" EML (@Jones2005).  This means that while figshare exposes only
a few of the metadata fields for search and discovery, a researcher can
search for particular attribute level information on the KNB.  Such as
"all data sets that have a column containing dissolved oxygen levels".

Publishing to the KNB is just as easy, after authentication credentials
have been set up.

```r
eml_publish(eml, permanent = TRUE, identifier = "doi", destination="knb")
```

The KNB provides several different kinds of identifiers, including
`uuid` and `doi`. A DOI (`doi`) should only be requested for finished,
published, public dataset, not individual files within a dataset or
in a dataset that is being regularly updated.  By default, the `EML`
package uses `uuid` for unique ids of component files and EML packages.
The option `permanent = TRUE` can be omitted for testing purposes.

The KNB already hosts several decades worth of EML metadata files from
around the world, including all data collected by Long Term Ecological
Research (LTER) sites.  Researchers can make these powerful search
queries directly from the R terminal, and use the `EML` package to parse
and explore the resulting hits.

@@

By default, the `EML` package has provided us with the liberal CC0
declaration for the data, consistent with major repositories such as
Dryad and Figshare.

```{r}
eml_get(eml, "rights")
```

White _et al._ discuss the motivation for CC0 as the recommended default
for scientific data, while also pointing to other license options.
We could have provided many other possible rights when writing the EML
by using the `intellectualRights` argument in an `eml` or `eml_write`
function call:

```{r}
eml_write(dat_with_metadata,
          contact = "Carl Boettiger <cboettig@ropensci.org>",  
          title = "A test EML file",
          intellectualRights = "Name of an extablished license or
          copyright declaration, Text string describing the rights to
          reuse this data, request for appropriate acknowledgment, etc")
```

[^1]: Disclaimer: CB served as a reviewer and KR as handling editor for White et al.
